Rapid and Accurate Global PTM Discovery (G-PTM-D) Using Post-Acquisition Spectral Calibration and Defined Mass Windows

### 2017-04-13 paper.tex
many

\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
\usepackage{multirow}
\usepackage{subcaption}

@@

Post-acquisition calibration is a useful tool that is shown to increase the number of confidently identified peptides, and enables qualitatively distinguishing modifications with similar mass.
Then, mass-window searches are contrasted with narrow-mass and open-mass searches.
The improvements in the performance of the enhanced G-PTM-D over the originally proposed G-PTM-D, that are due to using mass-window searches are shown.
Finally, the application of mass-window searches to the discovery of novel modifications is discussed.

@@

A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration, REF).
We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The increased mass accuracy of the recalibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

The spectral calibration procedure consists of two steps: Step 1: A standard tolerance (e.g. 10 ppm) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy (ref)).
Step 2:  The identifications from step 1 are used to extract peak matches from the spectra, and a calibration procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 1.

The average mass error went down from -3.67ppm to -0.15ppm in the jurkat dataset and from -3.32ppm for the mouse dataset.
The standard deviations decreased from 3.64ppm to 2.82ppm, and from 3.07ppm to 2.54ppm for the jurkat and mouse datasets respectively.
Figure 2 shows the clear improvement in the error for the withheld identified peptides, as a result of applying spectral calibration.

A 10 ppm narrow tolerance search produces matches with mass differences in ppm given in~\ref{fgr:figure1}.
The average and standard deviations along with 1, 2, and 3 standard deviation widths are given in Table~\ref{tbl:calib}.

\begin{table}
  \caption{Calibration Results: ppm Mass Differences}
  \label{tbl:calib}
  \begin{tabular}{llll}
    \hline
    & prior& after smooth & after rf  \\
    \hline
    stdev&	1.553	&1.039&	0.898 \\
	avg&	-3.148&	0.204&0.215\\
	68.3\% within	& [-4.701,-1.595 ]&	[-0.834,1.243	]&	[-0.683,1.114]\\
	95.4\% within&[	-6.254,	-0.042]&[	-1.873,2.282]&[	-1.581,	2.012]\\
	99.7\% within&	[-7.807,	1.511]&[	-2.912,3.321]&[	-2.479,	2.910]\\
    \hline
  \end{tabular}
\end{table}

\begin{figure}
  \includegraphics[scale=0.6]{figure_1.eps}
  \caption{An example figure}
  \label{fgr:figure1}
\end{figure}

\subsection{Mass-Window Searches}
A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da).
A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account.
We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches.
The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses.
The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches.
The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum).
Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.

>>>>

\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{booktabs}

@@

The first modification to the G-PTM-D workflow, post-acquisition calibration, is shown to increase the number of confidently identified peptides, and enables qualitatively distinguishing modifications with similar mass.
The other major enhancement, the mass-window search, is contrasted with the narrow-mass and open-mass search.
The superior performance of G-PTM-D achieved by the use of mass-window searches is demonstrated.
Finally, the application of mass-window searches to the discovery of novel modifications is discussed.

@@

A critical parameter for peptide identification and PTM localization is mass accuracy\cite{Scherl_2008}.
Higher mass accuracy provides increased specificity and thus confidence in peptide identifications, decreasing the false discovery rate.
Instrument noise, systemic drift and miscalibration limit the mass accuracy in acquired spectra.
Multiple calibration strategies to improve mass accuracy have been devised, and fall into three general categories:  External calibration prior to the MS experiment (e.g. standard instrument calibration protocols); internal calibration during the MS experiment (e.g. real-time calibration using a lock mass standard\cite{Olsen_2005}); and subsequent to the MS experiment (post-acquisition spectral calibration).
We use a post-acquisition calibration procedure, that builds upon the software lock mass concept\cite{Cox_2011} recently reported by the Mann group.
In our strategy, the m/z differences between expected and observed peaks in the peptide tandem ms spectra are compiled, and then used to recalibrate the spectra.
The increased mass accuracy of the calibrated spectra leads directly to improved identifications of both modified and unmodified peptides, as well as to increased confidence for PTM localization.

The spectral calibration procedure consists of two steps: Step 1: A narrow tolerance (e.g. 10 ppm precursor, 0.01 Dalton fragment) database search is performed on the dataset.
This yields a set of peptide identifications subject to a desired false discovery rate (e.g 1\%; based upon the target:decoy strategy \cite{Elias_2007}).
Step 2:  The identifications from Step 1 are used to extract peak matches from the spectra, and a calibration procedure is performed.
The calibration algorithm is described in detail in Supplementary Material 1.

Calibration quality is accessed through the results of a standard 10ppm precursor tolerance database search.
PSMs within 1\% FDR prior to calibration are compared with PSMs after the calibration procedure, see Table~\ref{tbl:calib}.
The calibration procedure successfully addresses instrument miscalibration by centering the average error around zero.
Furthermore, the systemic noise introduced by various other factors is addressed: this is evident by the decrease in the standard deviation of the errors in the precursor mass estimates, which places calibrated data squarely in the sub-ppm mode.
Histograms of the mass errors before and after calibration are plotted in Figure~\ref{fgr:figure1}.

\begin{table}[]
\centering
\caption{PSMs Before and After Calibration}
\label{tbl:calib}
\begin{tabular}{@{}llllll@{}}
                                                                                &        & \multicolumn{2}{l}{Mouse} & \multicolumn{2}{l}{Jurkat} \\ \cmidrule(l){3-6} 
                                                                                &        & Orig        & Calib       & Orig        & Calib        \\ \cmidrule(l){3-6} 
\multicolumn{2}{l}{PSM Count}                                                            &    152208       &        155617 &         217612   &   219172        \\
\multicolumn{2}{l}{Average Error (ppm)}                                                           &    -1.09         &       -0.16      &      -2.10      &        -0.16      \\
\multicolumn{2}{l}{St. Dev of Error}                                                     &     3.35        &      0.90       &       2.15      &        0.91      \\
\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}Confidence\\ Intervals\end{tabular}} & 68.3\% &   [-4.44,2.27]          &    [-1.06,0.74]          &    [-4.26,0.05]          &  [-1.08,0.75]            \\
                                                                                & 95.4\% &   [-7.80,5.62]          &    [-1.97,1.64]         &   [-6.41,2.20]          &      [-1.99,1.61]         \\
                                                                                & 99.7\% &   [-11.15,8.98]          &      [-2.87,2.54]        &    [-8.56,4.36]          &      [-2.90,2.58]         \\ \cmidrule(l){3-6} 
\end{tabular}
\end{table}

\begin{figure}
  \includegraphics[scale=0.85]{fig2-calibrationQuality.eps}
  \caption{Histograms demonstrating the effect of calibration on PSM mass errors}
  \label{fgr:figure1}
\end{figure}

\subsection{Mass-Window Searches}

A standard \textit{narrow-mass} search considers matching theoretical peptides to spectra if the peptide mass and the identified precursor mass are equal, within some tolerance (i.e. 5ppm, or 2.1 Da).
A \textit{wide-mass} search matches peptide fragment masses to fragmentation spectra while allowing large differences (i.e. $\pm 500$ Da) between the observed precursor mass and the peptide mass.
An even less restrictive \textit{open-mass} search, matches peptides from a database to fragmentation spectra without taking the precursor mass into account.
We propose alternatives to these search modes, namely a \textit{notch search} and an \textit{interval search} that fall under a broad category of \textit{mass-window} searches.
The notch search is an extension of the narrow-mass search: it allows exact matches (within a tolerance) to specific mass differences between the theoretical peptide and observed precursor masses.
The interval search is an extension of the wide-mass search: specific problematic mass differences are excluded from allowed matches.
The smaller the fraction of the mass space, and the narrower the mass window per search element, the faster the search will run (as fewer searches are executed per mass spectrum).
Figure~\ref{fgr:differentSearchModes} provides a convenient way to visualize the various search modes.
